NURM-tool: a quantitative reproducibility measurement tool to identify numerical differences in pipelines

### 2019-05-05 paper.tex

\begin{abstract}
Many experiments show that computational analysis results are still not 
completely reproducible. In particular, computational environments 
including different operating systems, hardware categories, and 
software versions are known to have effects on the results produced by 
analysis pipelines. These effects are presumably due to the creation, 
propagation and amplification of small numerical differences across the 
pipelines. 
Although new techniques such as virtualization , version 
control systems , and provenance management tools provide more reliable 
environment and significantly improve the reproducibility of scientific 
findings. 
However, the precise causes of such instabilities and the path along 
which they propagate in the pipelines are unclear.  We present a 
technique to identify the processes in the pipeline that create 
numerical differences along the execution, and we apply this technique to 
the HCP structural pre-processing pipelines.

@@

A reproducible study provides a context in which one is able to get 
results that are consistent with the original 
work~\cite{plesser2018reproducibility}. Research findings are expected 
to be reproducible so that their authenticity and reliability can be 
evaluated. 
There are different definitions for the term reproducibility, we follow 
Peng's definition~\cite{peng2011reproducible} in which reproducibility 
is defined as the ability to regenerate the same results as the 
original findings when the experiment is reanalyzed given exactly the 
same analytic methods, software package, parameters, and data.

In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. Two files will be considered numerically 
reproducible if they have identical binary contents. Binary comparison 
is calculated by comparing checksums. It must be pointed that a 
computation might be reproducible based on Peng's definition, but not 
be numerically reproducible. For instance, small numerical differences 
created during the pipeline execution may hamper numerical 
reproducibility, but be negligible in the final results.

Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible. Results show that the 
variety of computational infrastructures including workstation types, 
parallelization methods, operating systems, and analysis packages are 
known to influence reproducibility because of the creation of small 
numerical differences~\cite{Gronenschild2012, diethelm2012limits, 
Glatard2015, bowring2018exploring}. The propagation and amplification 
of these slight differences by analysis pipelines may cause reproducibility 
issues. 
In this case, the analysis pipelines are said to be numerically 
unstable. Numerical instability is a characteristic of the pipelines 
which amplify small numerical differences and then hamper the 
reproducibility of the analyses depending on the length of the pipeline 
and magnitude of the differences. In many cases, numerical instability is an 
important issue for reproducibility.

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} more specifically on some of 
the main neuroimaging pipelines including FSL, CIVET and Freesurfer 
pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating system, which indicate 
a significant disparity. 
The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported as the main influential 
computing infrastructure on reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques to encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NURM) tool to identify the processes in 
pipeline that create numerical differences across several runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using an interposition techniques.

The reminder of this paper describe the NURM-tool in different parts as 
the way of capturing provenance information and their representation, 
the influence of different type of subjects, and how to classify 
processes and characterize differences. We describe the HCP preprocessing 
pipelines as a popular neuroimaging pipelines to test and apply the 
NURM-tool on. Finally, we report the experimental results, the processes 
that are responsible for the differences in the pipelines.

@@

We present NURM-tool to automatically
measure numerical reproducibility of the pipelines across operating 
systems (see Figure~\ref{fig:overview}). We build the 
containerized version of the pipelines for each operating system using 
docker technology which encapsulated whole the pipeline dependencies. 
The docker images consists of an specific computing environment such as 
operating system version, hardware configuration, software version, 
etc. Furthermore, to simplify computational reproducibility of the 
analysis, we execute docker images using Boutiques framework based on 
the command-line descriptions~\cite{glatard2017boutiques}. The 
Boutiques descriptor specifies the parameters and implementation of the 
experiment, and the invocation schema describes the parameter values 
such as input/output data path.

As shown in the left side of Figure~\ref{fig:overview}, NURM-tool clusters 
the input subjects into various types based on the similarity of their 
process trees which are achieved by \reprozip tool. The right side of this figure 
shows the iterative classification of the process based on the binary 
comparison of the outputs from various conditions.
NURM-tool produces the \textit{difference} JSON schema in which 
recognizes output files with differences using the checksum comparison 
of the results. We consider that a file contains difference if only 
its binary comparison shows difference. Also, 
we use \reprozip tool to capture the provenance information. 
Subsequently, using a iterative approach, we classify processes into 
different categories regarding the state of read/write files 
with/without differences. NURM-tool identifies relevant processes in the 
pipeline creating, propagating, or removing differences. 
All the steps are explained in detail in the following sections.

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/overview.png}
  \caption{An overview of the proposed technique.}
  \label{fig:overview}
\end{figure}

@@

We use the \reprozip tool~\cite{Chirigati2016} to record: (1) 
the tree of processes executed by the pipeline and (2) the list of 
files read and written by each process. This information is collected 
by system call interceptions, through the \texttt{ptrace} Unix system 
call, and stored in a \texttt{SQLite} database. The database stores 
information on all the processes which are created by the 
\texttt{clone()} or \texttt{fork()} system calls. It also records the 
files opened by each process (in read, write or execution mode), 
including the result files and the temporary files.

NURM-tool reconstructs the tree of processes starting from the first process 
created by the pipeline and identifying its child processes as the ones 
that were created through \texttt{clone()} or \texttt{fork()}. It 
creates a process graph from this tree by adding edges corresponding to 
file dependencies between processes. A file dependency is defined 
between processes A and B if a file written by A is read by B. 
Figure~\ref{fig:simple_script} shows an example of a process tree 
constructed for the simple bash script illustrated in 
Algorithm~\ref{algo:sample-script} from the brain extraction procedure.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.3]{images/simple_graph}
  \caption{Process tree constructed from an example brain extraction
	script. Every node in the graph is labeled using (1) a process id 
	created by our reconstruction, (2) the name of the executable run 
	by the process. Processes that read or write temporary files are 
	represented with squares; other ones with circles. Plain edges 
	represent the process tree (\texttt{fork()} or \texttt{clone()} 
	system calls). Dashed edges represent file dependencies: temporary 
	files are in yellow and result files are in green.}
  \label{fig:simple_script}
\end{figure}

@@

Depending on the variety of computing environments, different  
process tree can be produced. In particular, different analysis 
software package can construct different process tree for the 
similar experiments because of following different methods. We used a 
fixed analysis software package in different conditions, however, we 
found that different subjects may produce a different process tree.

The challenging problem is that of comparing various process trees with 
different topological structures from different subjects. To address 
this issue, we categorize subjects into different types based on the 
distances, topologically differences, between corresponding process 
trees.

We use the tree edit distance~\cite{zhang1989simple} method to measure 
the distance between labeled trees. The distance between two labeled 
trees is defined as the minimum weighted number of edit operations that 
it takes to transform one tree to another. There are three edit 
operations that can be used to transform a tree including modify, 
remove, and insert. Modifying node \textit{n} means changing the label 
on \textit{n}. Removing a node \textit{n} means making the children of 
\textit{n} become the children of the parent of \textit{n} and then 
deleting \textit{n}. Inserting is the the complement of remove, this 
means that inserting \textit{n} as the child of \textit{n}$'$ will make 
\textit{n} the parent of a consecutive subsequence of the current 
children of \textit{n}$'$.
Each operation has an associated cost---insertions and removals each 
cost 1, and modifications cost either 1 or 2, depending on whether or 
not you consider a modification to essentially be just a removal 
followed by an insertion.

We cluster subjects into various types using the bottom-up hierarchical 
clustering approach implemented in \textit{scipy} python package (see 
Algorithm~\ref{also:hclustering}). This approach groups together the 
subjects with similar distance, less than a threshold.

\begin{algorithm}[h!]
\caption{Hierarchical clustering algorithm}
\label{algo:hclustering}
\begin{algorithmic}

  \State /* Input: a list of n trees; \texttt{T = [t${_1}$, t${_2}$,...,t${_n}$]}
  \State /* Output: a list of clusters; \texttt{C = [c${_1}$,c${_2}$,...,c${_k}$]}
  \For{\texttt{i=1} in \texttt{n}}
  \State \texttt{C${_i}$} = \texttt{T${_i}$} 
  \State \texttt{\# Create a cluster for each tree in input list; \texttt{agglomerative approach}}
  \EndFor
  \While{\texttt{len(C)} > \texttt{1}}  
  \For{\texttt{i=1} to \texttt{n}}
  \For{\texttt{j=i+1} to \texttt{n}}
  \State \texttt{D${_{ij}}$} = argmin\{\texttt{edit\_distance(C${_i}$, C${_j}$)}\}
  \State \texttt{\# Select two nearest clusters based on \texttt{nearest\_neighbors} method}
  \EndFor
  \EndFor
  \If{\texttt{D${_{ij}}$} > \texttt{thre}}
  \State Return \texttt{C}
  \Else {\State Merge \texttt{C${_i}$ and C${_j}$}}
  \EndIf
  \EndWhile
%  \State Function edit\_distance (C${_1}$, C${_2}$)
%  \State EndFunction
\end{algorithmic}
\end{algorithm}

The clustering algorithm starts by finding the two subjects that are 
closest to each other on the basis of edit distance method and form the 
new cluster between these two subjects. The next step join this 
cluster to the next nearest cluster or subject which in turn results in 
another cluster. This process continues until all the subjects are 
joined together to form one big cluster or terminates when the distances 
are less than the threshold. The threshold is defined as the minimum 
distance required to separate clusters.

@@

\begin{figure}
  \includegraphics[width=1\textwidth]{images/graph}
  \caption{A complete process graph from the PreFreesurfer pipeline.
Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
  \label{fig:complete-graph}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{images/hclusters}
  \caption{Different data types clustered among 100 subjects.}
  \label{fig:subj-clusters}
\end{figure}

>>>>

\begin{abstract}

Many experiments show that computational analysis results are still not 
completely reproducible. Computational environments including different 
operating systems, hardware categories, and software versions are known 
to have effects on the results produced by analysis pipelines. These 
effects are presumably due to the creation, propagation and 
amplification of small numerical differences across the pipelines. 
Although new techniques such as virtualization, version control 
systems, and provenance management tools provide more reliable 
environment and significantly improve the reproducibility of scientific 
findings. However, the precise causes of such instabilities and the 
path along which they propagate in the pipelines are unclear.  We 
present a technique to identify the processes in the pipeline that 
create numerical differences along the execution, and we apply this 
technique to the HCP structural pre-processing pipelines.

@@

A reproducible study provides a context in which one is able to get 
results that are consistent with the original 
work~\cite{plesser2018reproducibility}. Research findings are expected 
to be reproducible so that their authenticity and reliability can be 
evaluated.

Reproducibility is defined as the ability to regenerate the same 
results as the original findings when the experiment is reanalyzed by 
exactly the same analytic methods, software package, parameters, and 
data~\cite{peng2011reproducible}.

In addition, numerical reproducibility is defined as the ability to 
regenerate bit for bit identical results from multiple 
runs~\cite{hill2017numerical}. This Numerical reproducibility can be 
checked by comparing the binary content of the results which is 
computed by the checksum method.
Regarding these definitions, it must be pointed that a
computation might be reproducible, but not 
be numerically reproducible. For instance, small numerical differences 
created during the pipeline execution may hamper numerical 
reproducibility, but be negligible in the final results.

Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible. According to the previous 
studies, variety of computational infrastructures including workstation 
types, parallelization methods, operating systems, and analysis 
packages are known to influence reproducibility~\cite{Gronenschild2012, 
diethelm2012limits, Glatard2015, bowring2019exploring}.

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} specially on some of the main 
neuroimaging pipelines including FSL, CIVET and Freesurfer pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating systems, which 
indicate a significant disparity.

This irreproducibility issues are reported as the result of the 
creation, propagation, and amplification of small numerical 
differences~\cite{Gronenschild2012, diethelm2012limits, Glatard2015, 
bowring2019exploring}. In this case, the analysis pipelines are said to 
be numerically unstable. Numerical instability is a characteristic of 
the pipelines which amplify small numerical differences and then hamper 
the reproducibility of the analyses depending on the length of the 
pipeline and magnitude of the differences. In many cases, numerical 
instability is an important issue for reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques that encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NURM) tool to identify the processes in 
pipeline that create numerical differences across several runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using the interposition techniques.

The remainder of this paper describes the NURM-tool in different parts 
including the way of capturing and representation of the provenance 
information, the method of clustering of different subject types, and 
process classification to characterize differences. We describe the HCP 
preprocessing pipelines as a popular neuroimaging project to test the 
NURM-tool and evaluate its functionality. Finally, we reported the 
experimental results which describe the processes that are responsible 
for the differences in the pipelines.

@@

We present NURM-tool to automatically measure numerical reproducibility 
of the pipelines across operating systems (see 
Figures~\ref{fig:overview-cluster}, \ref{fig:overview-classify}). We build the 
containerized version of the pipelines for each operating system using 
docker technology which encapsulated whole the pipeline dependencies. 
The docker images consists of an specific computing environment such as 
operating system version, hardware configuration, software version, 
etc. Furthermore, to simplify computational reproducibility of the 
analysis, we execute docker images using Boutiques framework based on 
the command-line descriptions~\cite{glatard2017boutiques}. The 
Boutiques descriptor specifies the parameters and implementation of the 
experiment, and the invocation schema describes the parameter values 
such as input/output data path.

As shown in Figure~\ref{fig:overview-cluster}, NURM-tool clusters the 
input subjects into various types based on the similarity of their 
process trees which are achieved by \reprozip tool. 
Figure~\ref{fig:overview-classify} shows the iterative classification 
of the process based on the binary comparison of the outputs from 
various conditions. NURM-tool produces the \textit{difference} JSON 
schema in which recognizes output files with differences using the 
checksum comparison of the results. We consider that a file contains 
difference if only its binary comparison shows difference. Also, we use 
\reprozip tool to capture the provenance information. Subsequently, 
using a iterative approach, we classify processes into different 
categories regarding the state of read/write files with/without 
differences. NURM-tool identifies relevant processes in the pipeline 
creating, propagating, or removing differences. All the steps are 
explained in detail in the following sections.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{images/Slide1}
  \caption{An overview of the subject type clustering method.}
  \label{fig:overview-cluster}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.75\columnwidth]{images/Slide2}
  \caption{An overview of the process classification method.}
  \label{fig:overview-classify}
\end{figure}

%\begin{figure*}
%\centering
%    \begin{subfigure}{\columnwidth}
%        \includegraphics[width=\textwidth]{images/Slide1.png}
%        \caption{Subject clustering}
%        \label{fig:green}
%    \end{subfigure}
%\hfill%\hfil
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\textwidth]{images/Slide2.png}
%    \caption{Process classification}
%    \label{fig:red}
%\end{subfigure}
%    \caption{An overview of the proposed technique.}
%    \label{fig:processes}
%\end{figure*}

@@

We use the \reprozip tool~\cite{Chirigati2016} to record: (1) 
the tree of processes executed by the pipeline and (2) the list of 
files read and written by each process. This information is collected 
by system call interceptions, through the \texttt{ptrace} Unix system 
call, and stored in a \texttt{SQLite} database. The database stores 
information of all the processes which are created by the 
\texttt{clone()} or \texttt{fork()} system calls. It also records the 
files opened by each process (in read, write or execution mode), 
including the result files and the temporary files.

NURM-tool reconstructs the tree of processes starting from the first 
process created by the pipeline and through identifying its child 
processes as the ones that were created through \texttt{clone()} or 
\texttt{fork()}. It creates a process graph from this tree by adding 
edges corresponding to file dependencies between processes. A file 
dependency is defined between processes A and B if a file written by A 
is read by B. Figure~\ref{fig:simple_script} shows an example of a 
process tree constructed for the simple bash script illustrated in 
Algorithm~\ref{algo:sample-script} from the brain extraction procedure.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.25]{images/simple_graph}
  \caption{Process tree constructed from an example brain extraction
	script. Every node in the graph is labeled using (1) a process id 
	created by our reconstruction, (2) the name of the executable run 
	by the process. Processes that read or write temporary files are 
	represented with squares; other ones with circles. Plain edges 
	represent the process tree (\texttt{fork()} or \texttt{clone()} 
	system calls). Dashed edges represent file dependencies: temporary 
	files are in yellow and result files are in green.}
  \label{fig:simple_script}
\end{figure}

@@

Depending on the variety of computing environments, in particular, 
different analysis software packages, different process trees can be 
constructed for the same experiment. It was expected to achieve the 
same process trees since we used the same computing environments 
specially the same analysis software package in different conditions. 
However, we found that different subjects may result in different 
process trees.

Therefore, all subjects should be grouped into different types based on 
the topological structure of the process trees.
To this order, the distance between trees, topological differences, 
need to be calculated.

We used the tree edit distance~\cite{zhang1989simple} method to measure 
the distance between labeled trees. The distance between two labeled 
trees is defined as the minimum weighted number of edit operations that 
it takes to transform one tree to another. There are three edit 
operations that can be used to transform a tree including modify, 
remove, and insert. Modifying node \textit{n} means changing the label 
on \textit{n}. Removing node \textit{n} means making the children of 
\textit{n} become the children of its parent and then 
deleting it. Inserting is the complement of remove; this 
means that inserting \textit{n} as the child of \textit{n}$'$ will make 
\textit{n} the parent of a consecutive subsequence of the current 
children of \textit{n}$'$.
Each operation has an associated cost---insertions and removals each 
cost 1, and modifications cost either 1 or 2, depending on whether 
you consider a modification to essentially be just a removal 
or followed by an insertion.

We clustered subjects into various types using the bottom-up hierarchical 
clustering approach implemented in \textit{scipy} python package (see 
Algorithm~\ref{algo:hclustering}). This approach groups together the 
subjects with similar distance, less than a threshold.

% \texttt{agglomerative approach}
%  based on \texttt{nearest\_neighbors} method

\begin{algorithm}[h!]
\caption{Hierarchical clustering algorithm from \textit{scipy}}
\label{algo:hclustering}
\begin{algorithmic}

  \State /* Input: a list of n trees; \texttt{T = [t${_1}$, t${_2}$,...,t${_n}$]}
  \State /* Output: a list of clusters; \texttt{C = [c${_1}$,c${_2}$,...,c${_k}$]}
  \For{\texttt{i=1} in \texttt{n}}
  \State \texttt{C${_i}$} = \texttt{T${_i}$} 
  \State \texttt{\# Create a cluster for each tree in input list;}
  \EndFor
  \While{\texttt{len(C)} > \texttt{1}}  
  \For{\texttt{i=1} to \texttt{n}}
  \For{\texttt{j=i+1} to \texttt{n}}
  \State \texttt{D${_{ij}}$} = \texttt{edit\_distance(C${_i}$, C${_j}$)}
  \State \texttt{\# Select two nearest clusters;}
  \EndFor
  \EndFor
  \State \texttt{d\_min${_{ij}}$} = \texttt{MIN(D)}
  \If{\texttt{d\_min${_{ij}}$} > \texttt{thre}}
  \State Return \texttt{C}
  \Else {
  \State \texttt{C${_i}$} = Merge(\texttt{C${_i}$ and C${_j}$})
  \State Remove \texttt{C${_j}$}}
  \EndIf
  \EndWhile
%  \State Function edit\_distance (C${_1}$, C${_2}$)
%  \State EndFunction
\end{algorithmic}
\end{algorithm}

The clustering algorithm starts by finding two closest subjects to each 
other on the basis of edit distance method and form a new cluster 
between them. In the next iteration, this cluster aggregates with the 
next nearest cluster or subject and turns to another cluster. This 
process continues until all the subjects are aggregated to form one big 
cluster or the distances are less than an specified threshold. The 
threshold is defined as the minimum distance required to separate 
clusters.

@@

\begin{figure*}
\centering
  \includegraphics[width=.9\textwidth]{images/graph}
  \caption{A complete process graph from the PreFreesurfer pipeline.
Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
  \label{fig:complete-graph}
\end{figure*}

\begin{figure*}
\centering
  \includegraphics[width=.8\textwidth]{images/hclusters}
  \caption{Different data types clustered among 100 subjects.}
  \label{fig:subj-clusters}
\end{figure*}
